---
title: "Multiple Sclerosis (MS): Transcriptomic Changes After Treatment"
author: "Aida Davoudian Beni, Sara Ashna"
date: "2026-05-07"
output: 
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: cosmo
    highlight: tango
---

```{r setup, include=FALSE}
# Global chunk options: show code, suppress warnings/messages, set figure size
knitr::opts_chunk$set(
  echo    = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width  = 10,
  fig.height = 6
)
```

# Introduction

This analysis investigates transcriptomic differences between pre-treatment and
post-treatment samples from Multiple Sclerosis (MS) patients using the RNA-seq
dataset **SRP045500**. The pipeline covers:

1. Data loading and metadata preparation
2. Exploratory data analysis and quality control
3. Gene filtering
4. Normalization and GC-bias correction
5. Principal component analysis (PCA)
6. Differential expression analysis (DESeq2 and edgeR)
7. Heatmap of top shared DEGs
8. Functional enrichment analysis (ORA and GSEA)

---
```{r}
# Create figures directory
if (!dir.exists("figures")) {
  dir.create("figures")
}
```

# Library Loading

```{r libraries}
# Core Bioconductor packages
library(DESeq2)
library(SummarizedExperiment)
library(EDASeq)
library(edgeR)
library(biomaRt)
library(limma)
library(pheatmap)
library(clusterProfiler)
library(org.Hs.eg.db)
library(ReactomePA)
library(enrichplot)

# Tidyverse and utility packages
library(dplyr)
library(stringr)
library(ggplot2)
library(gridExtra)
library(plyr)
library(corrplot)
library(ggfortify)
library(ggrepel)
```

---

# Data Loading and Metadata Preparation

```{r data-loading, results='hide'}
# Load the SRP045500 SummarizedExperiment object
data <- readRDS("/opt/omicsdata/datasets/SRP045500.rds")

# Inspect the metadata column names and sample attributes
colnames(colData(data))
attr <- colData(data)$sra.sample_attributes

# Count pre- and post-treatment samples
sum(grepl("posttreatment", attr))
sum(grepl("pretreatment",  attr))

# Assign treatment labels based on attribute strings
treatment <- ifelse(
  grepl("posttreatment", attr), "post",
  ifelse(grepl("pretreatment", attr), "pre", NA)
)

# Extract cell type information from sample attributes
colData(data)$celltype <- sub(".*celltype;;([^|;]+).*", "\\1", attr)

# Subset to pre- and post-treatment samples only (remove NAs)
keep       <- !is.na(treatment)
data_ms    <- data[, keep]
treatment_ms <- treatment[keep]

# Store treatment labels in the SummarizedExperiment metadata
colData(data_ms)$treatment <- treatment_ms
```

---

# Exploratory Data Analysis and Quality Control

## Raw Count Distribution

```{r raw-count-boxplot}
# Extract raw count matrix
raw_count <- assay(data_ms, "raw_counts")

# Colour samples by treatment group
col <- ifelse(treatment_ms == "post", "purple", "steelblue")

# Boxplot of log2-transformed raw counts — shows per-sample expression spread
boxplot(
  log2(raw_count + 1),
  col      = col,
  las      = 2,
  cex.axis = 0.5,
  ylab     = "log2(counts + 1)",
  main     = "Raw count distribution (log2 scale)"
)
legend("topright", legend = c("post", "pre"), fill = c("purple", "steelblue"))
ggsave("figures/01_raw_counts_boxplot.png", width = 12, height = 6, dpi = 300)
```

## MA Plot (Raw Counts)

```{r ma-plot}
# MA plot: M = log2(sample / virtual reference), A = average log expression
# Ideally, most genes should centre around M = 0 (no overall fold change)
limma::plotMA(
  log1p(raw_count),
  main = "MA Plot — MS raw counts",
  xlab = "Average log expression",
  ylab = "M = log2(sample / reference)"
)
ggsave("figures/MA_Plot.png", width = 12, height = 6, dpi = 300)

```

## MD Plot (One Post vs One Pre Sample)

```{r md-plot}
# Compare the first post-treatment sample against the first pre-treatment sample
post_idx <- which(treatment_ms == "post")[1]
pre_idx  <- which(treatment_ms == "pre")[1]

MDPlot(raw_count, c(post_idx, pre_idx))
title(main = "MD Plot: one post vs one pre sample")
ggsave("figures/MD_Plot.png", width = 12, height = 6, dpi = 300)

```

---

# Gene Filtering

```{r filtering}
# Dimensions before filtering
cat("Dimensions before filtering:\n")
dim(data_ms)

# Remove lowly expressed genes: keep genes with total counts >= 10
data_ms_filt <- data_ms[rowSums(assay(data_ms)) >= 10, ]
raw_count_ms <- assay(data_ms_filt)   # filtered count matrix

# Dimensions after filtering
cat("\nDimensions after filtering:\n")
dim(data_ms_filt)
cat("Genes removed:", nrow(data_ms) - nrow(data_ms_filt), "\n")
```

---

# Principal Component Analysis (PCA) — Raw Counts

```{r pca-raw-summary}
# PCA on log-transformed raw counts (unfiltered) to assess overall structure
pca <- prcomp(t(log1p(raw_count)))

# Variance explained by the first 20 PCs
cat("Variance explained by first 20 PCs:\n")
print(summary(pca)$importance[, 1:20])
```

```{r pca-vst}
# Build DESeq2 object from FILTERED data for VST-based PCA
dds <- DESeqDataSet(data_ms_filt, design = ~ treatment)
dds$treatment <- relevel(factor(dds$treatment), ref = "pre")

# Variance Stabilising Transformation (blind = TRUE for exploratory QC)
vsd <- vst(dds, blind = TRUE)

# PCA on VST-transformed data; samples are rows after transposition
pca_full    <- prcomp(t(assay(vsd)))
pca_df      <- as.data.frame(pca_full$x[, 1:4])
pca_df$treatment <- treatment_ms

# Percentage of variance explained by each PC
percentVar_full <- round(100 * summary(pca_full)$importance[2, 1:4], 1)

# Screeplot — helps decide how many PCs capture meaningful variation
screeplot(pca_full, type = "lines", main = "Variance explained by each PC")

# Three pairwise PC comparison plots
p1 <- ggplot(pca_df, aes(PC1, PC2, color = treatment)) +
  geom_point(size = 3) + stat_ellipse(type = "t", level = 0.95) +
  scale_color_manual(values = c("post" = "purple", "pre" = "steelblue")) +
  xlab(paste0("PC1: ", percentVar_full[1], "% variance")) +
  ylab(paste0("PC2: ", percentVar_full[2], "% variance")) +
  ggtitle("PC1 vs PC2") + theme_bw()

p2 <- ggplot(pca_df, aes(PC1, PC3, color = treatment)) +
  geom_point(size = 3) + stat_ellipse(type = "t", level = 0.95) +
  scale_color_manual(values = c("post" = "purple", "pre" = "steelblue")) +
  xlab(paste0("PC1: ", percentVar_full[1], "% variance")) +
  ylab(paste0("PC3: ", percentVar_full[3], "% variance")) +
  ggtitle("PC1 vs PC3") + theme_bw()

p3 <- ggplot(pca_df, aes(PC2, PC3, color = treatment)) +
  geom_point(size = 3) + stat_ellipse(type = "t", level = 0.95) +
  scale_color_manual(values = c("post" = "purple", "pre" = "steelblue")) +
  xlab(paste0("PC2: ", percentVar_full[2], "% variance")) +
  ylab(paste0("PC3: ", percentVar_full[3], "% variance")) +
  ggtitle("PC2 vs PC3") + theme_bw()

grid.arrange(p1, p2, p3, ncol = 3)
# 02-04 - Individual PCA plots
ggsave("figures/02_pca_pc1_pc2.png", p1, width = 8, height = 6, dpi = 300)
ggsave("figures/03_pca_pc1_pc3.png", p2, width = 8, height = 6, dpi = 300)
ggsave("figures/04_pca_pc2_pc3.png", p3, width = 8, height = 6, dpi = 300)

# 05 - Combined PCA
combined_pca <- grid.arrange(p1, p2, p3, ncol = 3)
ggsave("figures/05_pca_combined.png", combined_pca, width = 18, height = 6, dpi = 300)
```

---

# Normalization and Bias Correction

## RLE Plot (Raw Counts)

```{r rle-raw}
# Relative Log Expression (RLE) plot on raw counts.
# Boxes should ideally be centred at 0; any offset indicates normalization need.
plotRLE(
  raw_count_ms,
  outline = FALSE, las = 2,
  ylab = "RLE", main = "RLE of raw counts", col = col
)
```

## GC-Content Bias Correction (EDASeq)

```{r gc-bias-correction}
# Connect to Ensembl and retrieve GC content and gene length for all genes
mart         <- useMart("ensembl", dataset = "hsapiens_gene_ensembl")
gene_ids_original <- rownames(raw_count_ms)
gene_ids_clean    <- sub("\\..*", "", gene_ids_original)   # strip version suffixes

gene_info <- getBM(
  attributes = c("ensembl_gene_id", "percentage_gene_gc_content",
                 "start_position", "end_position"),
  filters    = "ensembl_gene_id",
  values     = gene_ids_clean,
  mart       = mart
)
gene_info$length <- gene_info$end_position - gene_info$start_position

# Remove duplicate Ensembl IDs and align to count matrix
gene_info_unique <- gene_info[!duplicated(gene_info$ensembl_gene_id), ]
found_in_ensembl <- gene_ids_clean %in% gene_info_unique$ensembl_gene_id
counts_gc        <- raw_count_ms[found_in_ensembl, ]
kept_clean_ids   <- gene_ids_clean[found_in_ensembl]

keep_unique          <- !duplicated(kept_clean_ids)
counts_gc_dedup      <- counts_gc[keep_unique, ]
kept_clean_ids_dedup <- kept_clean_ids[keep_unique]
match_idx_dedup      <- match(kept_clean_ids_dedup, gene_info_unique$ensembl_gene_id)
gene_info_dedup      <- gene_info_unique[match_idx_dedup, ]
rownames(counts_gc_dedup) <- kept_clean_ids_dedup

# Build SeqExpressionSet with GC content (as percentage 0–100) and gene length
featuredata <- data.frame(
  gc     = gene_info_dedup$percentage_gene_gc_content,
  length = gene_info_dedup$length,
  row.names = kept_clean_ids_dedup
)

phenodata <- data.frame(
  treatment = treatment_ms,
  row.names = colnames(counts_gc_dedup)
)

set_ms <- newSeqExpressionSet(
  counts      = as.matrix(counts_gc_dedup),
  featureData = AnnotatedDataFrame(featuredata),
  phenoData   = AnnotatedDataFrame(phenodata)
)

# Drop genes with missing GC content
set_ms <- set_ms[!is.na(fData(set_ms)$gc), ]

# GC bias plot BEFORE normalization
colbias <- ifelse(treatment_ms == "post", "purple", "steelblue")
biasPlot(set_ms, "gc", log = TRUE, ylim = c(0, 20),
         col = colbias, main = "GC bias BEFORE normalization")

# withinLaneNormalization requires GC as a proportion (0–1)
fData(set_ms)$gc <- fData(set_ms)$gc / 100
wt_ms <- withinLaneNormalization(set_ms, "gc", which = "full", offset = FALSE)

# Convert GC back to percentage for the post-normalization bias plot
fData(wt_ms)$gc <- fData(wt_ms)$gc * 100
biasPlot(wt_ms, "gc", log = TRUE, ylim = c(0, 20),
         col = colbias, main = "GC bias AFTER normalization")
```

## Between-Sample Normalization

```{r between-sample-norm}
# Between-lane (between-sample) normalization using EDASeq
set_full <- betweenLaneNormalization(wt_ms, which = "full", offset = FALSE)

# Extract the within-lane normalized counts for edgeR comparisons
raw_count_norm <- counts(wt_ms)

# Apply additional edgeR normalization methods for comparison
nf_tmm <- calcNormFactors(raw_count_norm, method = "TMM")
tmm    <- cpm(raw_count_norm, lib.size = nf_tmm * colSums(raw_count_ms))

nf_rle <- calcNormFactors(raw_count_norm, method = "RLE")
rle    <- cpm(raw_count_norm, lib.size = nf_rle * colSums(raw_count_ms))

nf_uq  <- calcNormFactors(raw_count_norm, method = "upperquartile")
uq     <- cpm(raw_count_norm, lib.size = nf_uq  * colSums(raw_count_ms))

# Total count normalization (library-size CPM)
tc <- cpm(raw_count_norm)
```

## Boxplots: Comparing Normalization Methods

```{r norm-boxplots, fig.height=8}
# Side-by-side log2 boxplots for each normalization method.
# Well-normalized samples should show aligned medians.
par(mfrow = c(2, 3), mar = c(3, 4, 3, 1))
ylim_shared <- c(0, 10)

boxplot(log2(raw_count_norm + 1), outline = FALSE, las = 2, col = col,
        ylab = "log2(count+1)", main = "Before normalization",
        xaxt = "n", ylim = ylim_shared)

boxplot(log2(tc + 1), outline = FALSE, las = 2, col = col,
        ylab = "log2(count+1)", main = "Total count",
        xaxt = "n", ylim = ylim_shared)

boxplot(log2(rle + 1), outline = FALSE, las = 2, col = col,
        ylab = "log2(count+1)", main = "RLE",
        xaxt = "n", ylim = ylim_shared)

boxplot(log2(tmm + 1), outline = FALSE, las = 2, col = col,
        ylab = "log2(count+1)", main = "TMM",
        xaxt = "n", ylim = ylim_shared)

boxplot(log2(uq + 1), outline = FALSE, las = 2, col = col,
        ylab = "log2(count+1)", main = "Upper quartile",
        xaxt = "n", ylim = ylim_shared)

par(mfrow = c(1, 1))
legend("topright", legend = c("post", "pre"),
       fill = c("purple", "steelblue"), border = NA, bty = "n", cex = 0.9)
```

## RLE Plots After Normalization

```{r rle-post-norm}
# Compare RLE spread before and after normalization.
# Good normalization compresses boxes toward zero with consistent spread.
par(mfrow = c(1, 3))

EDASeq::plotRLE(raw_count_norm, outline = FALSE, las = 2, col = col,
                ylab = "RLE", main = "RLE — Raw", xaxt = "n")

EDASeq::plotRLE(tmm, outline = FALSE, las = 2, col = col,
                ylab = "RLE", main = "RLE — TMM", xaxt = "n")

EDASeq::plotRLE(rle, outline = FALSE, las = 2, col = col,
                ylab = "RLE", main = "RLE — RLE norm.", xaxt = "n")

par(mfrow = c(1, 1))
```

## Sample Correlation Heatmap (TMM)

```{r correlation-plot}
# Pearson correlation matrix on log-TMM values, clustered by similarity.
# High intra-group correlation (post-post, pre-pre) is expected.
colnames_orig <- colnames(tmm)
colnames(tmm) <- substr(colnames(tmm), 9, 12)   # shorten labels for readability
m <- cor(log1p(tmm))
corrplot(m, method = "color", order = "hclust", tl.cex = 0.2,
         main = "Sample correlation — TMM")
colnames(tmm) <- colnames_orig   # restore original column names
```

## PCA After Normalization

```{r pca-norm}
# PCA on three normalization strategies to assess separation of treatment groups
col_num <- ifelse(treatment_ms == "post", 1, 2)
par(mfrow = c(1, 3))
EDASeq::plotPCA(tc,  k = 2, labels = FALSE, col = col_num, pch = 20, main = "PCA — TC")
EDASeq::plotPCA(tmm, k = 2, labels = FALSE, col = col_num, pch = 20, main = "PCA — TMM")
EDASeq::plotPCA(rle, k = 2, labels = FALSE, col = col_num, pch = 20, main = "PCA — RLE")
par(mfrow = c(1, 1))
legend("topright", legend = c("post", "pre"), col = c(1, 2), pch = 20)
```

---

# Differential Expression Analysis

## DESeq2

```{r deseq2}
# Build DESeq2 object from filtered raw counts, controlling for cell type
data_ms$treatment <- factor(data_ms$treatment, levels = c("pre", "post"))

dds <- DESeq2::DESeqDataSetFromMatrix(
  countData = raw_count_ms,
  colData   = colData(data_ms),
  design    = ~ celltype + treatment
)

# Full DESeq2 pipeline:
#   1. estimateSizeFactors  — depth normalization
#   2. estimateDispersions  — gene-wise variance estimation
#   3. nbinomWaldTest       — Wald test for significance
dds <- DESeq(dds)

# Extract results (post vs pre; pre is the reference level)
res    <- results(dds)
res_df <- as.data.frame(res)
res_df$gene <- rownames(res_df)

# Classify genes by significance and direction
res_df$significance <- "Not significant"
res_df$significance[res_df$padj < 0.05 & res_df$log2FoldChange >  1] <- "Upregulated"
res_df$significance[res_df$padj < 0.05 & res_df$log2FoldChange < -1] <- "Downregulated"

cat("DESeq2 results (FDR < 0.05, |log2FC| > 1):\n")
cat("  Upregulated:  ", sum(res_df$significance == "Upregulated"),   "\n")
cat("  Downregulated:", sum(res_df$significance == "Downregulated"), "\n")
```

## edgeR

```{r edger}
# Build edgeR DGEList from filtered raw counts
y <- DGEList(raw_count_ms, group = data_ms$treatment)

# Remove lowly expressed genes using the edgeR criterion
keep_edger <- filterByExpr(y, group = data_ms$treatment)
y <- y[keep_edger, , keep.lib.sizes = FALSE]

# TMM normalization
y <- calcNormFactors(y, method = "TMM")

# Design matrix controlling for cell type (same covariates as DESeq2)
design <- model.matrix(~ celltype + treatment, data = colData(data_ms))

# Estimate dispersions and fit quasi-likelihood model
y   <- estimateDisp(y, design)
fit <- glmQLFit(y, design)

# Test the treatment coefficient (post vs pre)
qlf <- glmQLFTest(fit, coef = "treatmentpost")

# Extract all results
edgeR_res       <- topTags(qlf, n = Inf)$table
edgeR_res$gene  <- rownames(edgeR_res)

# Classify genes by significance and direction
edgeR_res$significance <- "Not significant"
edgeR_res$significance[edgeR_res$FDR < 0.05 & edgeR_res$logFC >  1] <- "Upregulated"
edgeR_res$significance[edgeR_res$FDR < 0.05 & edgeR_res$logFC < -1] <- "Downregulated"

cat("edgeR results (FDR < 0.05, |logFC| > 1):\n")
cat("  Upregulated:  ", sum(edgeR_res$significance == "Upregulated"),   "\n")
cat("  Downregulated:", sum(edgeR_res$significance == "Downregulated"), "\n")

# Save edgeR results to file
write.csv(edgeR_res, "edgeR_results.csv", row.names = FALSE)
```

DESeq2 identified 256 upregulated and 67 downregulated genes (FDR < 0.05,
|log2FC| > 1) in post- vs pre-treatment MS samples, while edgeR identified 163
upregulated and 0 downregulated genes under the same thresholds. The higher
sensitivity of DESeq2 is consistent with its variance shrinkage approach.
Genes significant in **both** methods are treated as the most robust DEGs for
downstream analysis.

## DESeq2–edgeR Overlap

```{r overlap}
# Identify genes significant in DESeq2 and/or edgeR
deseq2_sig  <- res_df$gene[res_df$significance != "Not significant"]
edger_sig   <- edgeR_res$gene[edgeR_res$significance != "Not significant"]

shared_genes <- intersect(deseq2_sig, edger_sig)
deseq2_only  <- setdiff(deseq2_sig, edger_sig)
edger_only   <- setdiff(edger_sig, deseq2_sig)

cat("Shared genes:  ", length(shared_genes), "\n")
cat("DESeq2 only:   ", length(deseq2_only),  "\n")
cat("edgeR only:    ", length(edger_only),   "\n")

# Save gene lists
write.csv(data.frame(gene = shared_genes), "shared_genes.csv",      row.names = FALSE)
write.csv(data.frame(gene = deseq2_only),  "deseq2_only_genes.csv", row.names = FALSE)
write.csv(data.frame(gene = edger_only),   "edger_only_genes.csv",  row.names = FALSE)
```

## Volcano Plots

```{r volcano-plots}
# DESeq2 volcano: log2FC on x, -log10(adjusted p-value) on y
ggplot(res_df, aes(x = log2FoldChange, y = -log10(padj), color = significance)) +
  geom_point(alpha = 0.6, size = 1.5) +
  scale_color_manual(values = c("Upregulated"   = "purple",
                                "Downregulated" = "steelblue",
                                "Not significant" = "grey70")) +
  geom_vline(xintercept = c(-1, 1), linetype = "dashed") +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed") +
  labs(title = "Volcano plot — DESeq2", x = "log2 fold change", y = "-log10(adj. p-value)") +
  theme_bw()

# edgeR volcano
ggplot(edgeR_res, aes(x = logFC, y = -log10(FDR), color = significance)) +
  geom_point(alpha = 0.6, size = 1.5) +
  scale_color_manual(values = c("Upregulated"   = "purple",
                                "Downregulated" = "steelblue",
                                "Not significant" = "grey70")) +
  geom_vline(xintercept = c(-1, 1), linetype = "dashed") +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed") +
  labs(title = "Volcano plot — edgeR", x = "logFC", y = "-log10(FDR)") +
  theme_bw()

```

## Fold-Change Agreement (DESeq2 vs edgeR)

```{r fc-scatter}
# Merge results from both methods on the common gene column
merged2 <- merge(
  res_df[,    c("gene", "log2FoldChange", "padj", "significance")],
  edgeR_res[, c("gene", "logFC", "FDR", "significance")],
  by = "gene", suffixes = c("_DESeq2", "_edgeR")
)
colnames(merged2)[colnames(merged2) == "log2FoldChange"] <- "DESeq2_log2FC"
colnames(merged2)[colnames(merged2) == "logFC"]          <- "edgeR_logFC"

# Highlight genes significant in both methods
shared_only <- subset(merged2,
  significance_DESeq2 != "Not significant" &
  significance_edgeR  != "Not significant")

# Scatter plot: high concordance between methods is expected
ggplot(merged2, aes(x = DESeq2_log2FC, y = edgeR_logFC)) +
  geom_point(alpha = 0.4, size = 1.2, color = "grey50") +
  geom_point(data = shared_only, aes(x = DESeq2_log2FC, y = edgeR_logFC),
             color = "purple", alpha = 0.8, size = 1.5) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "red") +
  geom_smooth(method = "lm", se = FALSE, color = "blue", linewidth = 0.8) +
  labs(title = "Fold-change agreement: DESeq2 vs edgeR",
       x = "DESeq2 log2 fold change", y = "edgeR logFC") +
  theme_bw()
```

## Shared DEG Table

```{r shared-degs}
# Build a merged table retaining only genes significant in BOTH methods
deseq2_tbl <- res_df[, c("gene", "log2FoldChange", "padj", "significance")]
colnames(deseq2_tbl) <- c("gene", "DESeq2_log2FC", "DESeq2_padj", "DESeq2_class")

edger_tbl <- edgeR_res[, c("gene", "logFC", "FDR", "significance")]
colnames(edger_tbl) <- c("gene", "edgeR_logFC", "edgeR_FDR", "edgeR_class")

merged_results <- merge(deseq2_tbl, edger_tbl, by = "gene", all = FALSE)

shared_results <- merged_results[
  merged_results$DESeq2_class != "Not significant" &
  merged_results$edgeR_class  != "Not significant", ]

# Save merged and shared tables
write.csv(merged_results, "merged_DESeq2_edgeR_results.csv", row.names = FALSE)
write.csv(shared_results, "shared_DEGs_merged_table.csv",   row.names = FALSE)

head(shared_results)
```

```{r direction-check}
# Check whether fold-change directions agree between DESeq2 and edgeR
shared_results$agreement <- ifelse(
  sign(shared_results$DESeq2_log2FC) == sign(shared_results$edgeR_logFC),
  "Same direction", "Different direction"
)
table(shared_results$agreement)
```

---

# Heatmap of Top Shared DEGs

```{r heatmap}
# Select the top 40 shared DEGs by the smallest mean adjusted p-value
shared_results$mean_padj <- rowMeans(
  cbind(shared_results$DESeq2_padj, shared_results$edgeR_FDR), na.rm = TRUE
)
top_shared       <- shared_results[order(shared_results$mean_padj), ]
top_shared_genes <- head(top_shared$gene, 40)

# VST matrix from DESeq2 (blind = FALSE: uses design for dispersion estimation)
vsd <- vst(dds, blind = FALSE)
mat <- assay(vsd)[top_shared_genes, , drop = FALSE]

# Z-score each gene row (mean = 0, sd = 1) so colours reflect relative change
mat_scaled <- t(scale(t(mat)))

# Column annotation for treatment group
annotation_col <- data.frame(
  Treatment = factor(treatment_ms, levels = c("pre", "post"))
)
rownames(annotation_col) <- colnames(mat_scaled)

pheatmap(
  mat_scaled,
  annotation_col = annotation_col,
  show_colnames  = FALSE,
  fontsize_row   = 7,
  cluster_rows   = TRUE,
  cluster_cols   = TRUE,
  color          = colorRampPalette(c("steelblue", "white", "purple"))(100),
  main           = "Top 40 shared DEGs: DESeq2 and edgeR"
)
```

---

# Functional Enrichment Analysis

## Gene ID Mapping

```{r gene-id-mapping}
# Shared significant genes will be used as the foreground set for ORA
sign_genes    <- shared_results$gene
universe_genes <- merged_results$gene

# Strip Ensembl version suffixes (e.g. ENSG00000001234.5 -> ENSG00000001234)
sign_clean    <- sub("\\..*", "", sign_genes)
universe_clean <- sub("\\..*", "", universe_genes)

# Map Ensembl IDs to Entrez IDs (required by KEGG and ReactomePA)
sign.entrez <- mapIds(org.Hs.eg.db,
                      keys    = sign_clean,
                      column  = "ENTREZID",
                      keytype = "ENSEMBL",
                      multiVals = "first")

universe.entrez <- mapIds(org.Hs.eg.db,
                          keys    = universe_clean,
                          column  = "ENTREZID",
                          keytype = "ENSEMBL",
                          multiVals = "first")

# Remove NAs and deduplicate
sign.entrez     <- unique(as.character(sign.entrez[!is.na(sign.entrez)]))
universe.entrez <- unique(as.character(universe.entrez[!is.na(universe.entrez)]))
```

## Over-Representation Analysis (ORA)

### KEGG ORA

```{r kegg-ora}
# KEGG pathway over-representation test (BH-corrected, FDR < 0.05)
keg1 <- enrichKEGG(
  gene         = sign.entrez,
  universe     = universe.entrez,
  organism     = "hsa",
  keyType      = "ncbi-geneid",
  pvalueCutoff = 0.05,
  pAdjustMethod = "BH"
)

barplot(keg1, showCategory = 10, title = "KEGG ORA — barplot")
dotplot(keg1, showCategory = 10, title = "KEGG ORA — dotplot")
```

### Reactome ORA

```{r reactome-ora}
# Reactome pathway over-representation test
enrich_reactome <- enrichPathway(
  gene         = sign.entrez,
  universe     = universe.entrez,
  organism     = "human",
  pvalueCutoff = 0.05,
  pAdjustMethod = "BH"
)

barplot(enrich_reactome, showCategory = 10, title = "Reactome ORA — barplot")
dotplot(enrich_reactome, showCategory = 10, title = "Reactome ORA — dotplot")
```

### GO Biological Process ORA

```{r go-ora}
# GO Biological Process over-representation test
enrich_ego <- enrichGO(
  gene         = sign.entrez,
  universe     = universe.entrez,
  OrgDb        = org.Hs.eg.db,
  keyType      = "ENTREZID",
  ont          = "BP",
  pvalueCutoff = 0.05,
  pAdjustMethod = "BH",
  readable     = TRUE   # converts Entrez IDs to gene symbols in output
)

barplot(enrich_ego, showCategory = 10, title = "GO Biological Process ORA — barplot")
dotplot(enrich_ego, showCategory = 10, title = "GO Biological Process ORA — dotplot")
```

## Gene Set Enrichment Analysis (GSEA)

```{r gsea-ranklist}
# Build a ranked gene list from DESeq2 log2 fold changes (all genes, not just DEGs)
# Ranking by log2FC ensures the most strongly changed genes anchor each tail
deseq2_gsea            <- res_df[!is.na(res_df$log2FoldChange), ]
deseq2_gsea$gene_clean <- sub("\\..*", "", deseq2_gsea$gene)

deseq2_gsea$entrez <- mapIds(org.Hs.eg.db,
                             keys    = deseq2_gsea$gene_clean,
                             column  = "ENTREZID",
                             keytype = "ENSEMBL",
                             multiVals = "first")

# Remove unmapped and duplicated entries
deseq2_gsea <- deseq2_gsea[!is.na(deseq2_gsea$entrez), ]
deseq2_gsea <- deseq2_gsea[!duplicated(deseq2_gsea$entrez), ]

# Named, sorted vector of log2FC values (descending) for GSEA input
tt           <- deseq2_gsea$log2FoldChange
names(tt)    <- deseq2_gsea$entrez
tt           <- sort(tt, decreasing = TRUE)
```

### KEGG GSEA

```{r kegg-gsea}
kk_gsea <- gseKEGG(
  geneList      = tt,
  organism      = "hsa",
  keyType       = "ncbi-geneid",
  pvalueCutoff  = 0.05,
  pAdjustMethod = "BH",
  verbose       = FALSE
)

ridgeplot(kk_gsea, showCategory = 15, label_format = 50)
gseaplot(kk_gsea, geneSetID = 1, title = kk_gsea$Description[1])
```

### Reactome GSEA

```{r reactome-gsea}
reactome_gsea <- gsePathway(
  geneList      = tt,
  organism      = "human",
  pvalueCutoff  = 0.05,
  pAdjustMethod = "BH",
  verbose       = FALSE
)

dotplot(reactome_gsea, showCategory = 10, title = "GSEA — Reactome")
gseaplot(reactome_gsea, geneSetID = 1, title = reactome_gsea$Description[1])
```

### GO Biological Process GSEA

```{r go-gsea}
go_gsea <- gseGO(
  geneList      = tt,
  ont           = "BP",
  OrgDb         = org.Hs.eg.db,
  keyType       = "ENTREZID",
  pvalueCutoff  = 0.05,
  pAdjustMethod = "BH",
  verbose       = FALSE
)

dotplot(go_gsea, showCategory = 10, title = "GSEA — GO Biological Process")
gseaplot(go_gsea, geneSetID = 1, title = go_gsea$Description[1])
```

---

# Summary Tables of Top Enriched Terms

```{r top-terms}
# Extract the top 10 terms from each ORA method, ordered by adjusted p-value.
# These results are used to contextualise findings relative to the original paper.

kegg_top <- as.data.frame(keg1) %>%
  dplyr::arrange(p.adjust) %>%
  dplyr::select(ID, Description, GeneRatio, Count, p.adjust) %>%
  head(10)

reactome_top <- as.data.frame(enrich_reactome) %>%
  dplyr::arrange(p.adjust) %>%
  dplyr::select(ID, Description, GeneRatio, Count, p.adjust) %>%
  head(10)

go_top <- as.data.frame(enrich_ego) %>%
  dplyr::arrange(p.adjust) %>%
  dplyr::select(ID, Description, GeneRatio, Count, p.adjust) %>%
  head(10)

cat("=== Top 10 KEGG terms ===\n");    print(kegg_top)
cat("\n=== Top 10 Reactome terms ===\n"); print(reactome_top)
cat("\n=== Top 10 GO-BP terms ===\n");   print(go_top)
```

```{r combined-table}
# Combined table of all top enrichment results for comparison with the literature
all_top_terms <- bind_rows(
  mutate(kegg_top,     source = "KEGG"),
  mutate(reactome_top, source = "Reactome"),
  mutate(go_top,       source = "GO-BP")
)
print(all_top_terms)
```

---

*End of analysis.*
